Automatic Amortized Analysis for Imperative Code

### 2014-05-14 main.tex
typo
merge with 690 report

\usepackage{amsthm}
\theoremstyle{newstyle}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

@@

\section{Decreasing to zero}

We start with a programming language with loops counting
downwards to 0.  Statements and expressions are defined
below.  Assignments of complex expressions can be simulated
by sequencing assignments of simple expressions.
%
\begin{mathpar}

S := \code{skip}
\mid \code{assert}(\Gamma)
\mid x \gets E
\mid \code{while}(0 < x)~S
\mid \code{if}(?)~S~\code{else}~S
\mid S;S

@@

E := x + y \mid x - y % \mid y
\end{mathpar}
%
We define a deductive system to derive judgements of the
form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.  The semantics
of this judgement are defined below in the soundness section.

\begin{mathpar}

\Rule{Skip}
{}
{ \Gamma; Q + M_s \vdash \code{skip} \dashv Q; \Gamma }

\and \Rule{Assert}
{}
{ \Gamma; Q + M_a \vdash \code{assert}(\Delta) \dashv Q; \Gamma, \Delta }

% \\ \Rule{AssignP}
% { \Gamma \models \Gamma'[x/x+y] }
% {\Gamma; Q + M_s + y\cdot q_x \vdash x \gets x + y \dashv Q; \Gamma'}

% \and\Rule{AssignM}
% {  \Gamma \models \Gamma'[x/x-y]
% \\ \Gamma \models c \le y \le x
% }
% { \Gamma; Q + M_s \vdash x \gets x - y \dashv c \cdot q_x + V; \Gamma' }

\\ \Rule{While}
{ \Gamma, 0 < x ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
{ \Gamma; Q \vdash \code{while}(0 < x)~S \dashv Q - M_w^3; \Gamma, x \le 0 }

\and\Rule{Seq}
{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }

\end{mathpar}

\section{Decreasing differences}
The second iteration is a system more symmetric where the potentials
are linear functions in differences of variables.  From now on $k$
denotes a non-negative constant.
$$
S := \code{while}(k < x - y)~S \mid \dots
$$
%
In the rules below we take as convention that $q_{x}$ for a variable
$x$ is in fact $q_{x0}$ where 0 is a special variable fixed to zero
in all contraint contexts.

\begin{mathpar}

\Rule{AssignOpK}
{  \Gamma \models \Gamma'[x/x \pm y]
\\ \Gamma \models |y| = k
\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
\\ \textstyle q'_y = q_y
      + \sum_{u \in \mathcal U} p_{xu}
      - \sum_{v \not\in \mathcal U} q_{xv}
\\ q_{xu} = p_{xu} + r_{xu}
}
{ \textstyle
  \Gamma; Q + M_s
  \vdash x \gets x \pm y \dashv
  \sum_u k \cdot r_{xu} + Q'; \Gamma'
  % q_x is not omitted, v_{x} is in fact a shorthand for
  % v_{x0} where 0 is a variable equal to 0.
}

\and \Rule{AssignOp}
{  \Gamma \models \Gamma'[x/x \pm y]
\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
\\ \textstyle q'_y = q_y
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma'
}

\\ \mbox{(The other rules are unchanged)}

\end{mathpar}

\section{Cost aware semantics}

This big step semantics is parameterized by a metric that sets
the cost of each program operation.  Non-deterministically,
the rule {\sc S:Abort} can be called to interrupt a computation.
It has the same effect as raising an exception.  We symbolize a
frozen computation with the notation $\circ$.  In the rules, $X$
stands for either $\circ$ or $H$.

\begin{mathpar}

\Rule{S:Abort}
{ }
{ (H, S) \Downarrow_0 \circ }

\and \Rule{S:Skip}
{ }
{ (H, \code{skip}) \Downarrow_{M_s} H }

@@

\and \Rule{S:Assert}
{ H \models \Gamma }
{ (H, \code{assert}(\Gamma)) \Downarrow_{M_a} H }

\and \Rule{S:WhileStop}
{ \neg (k < \llbracket x - y \rrbracket_H) }
{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^3} H }

\and \Rule{S:WhileStep}
{  k < \llbracket x - y \rrbracket_H
\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
\\ (H', \code{while}(k < x - y)~S) \Downarrow_{(q_2, q'_2)} X
}
{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }

\and \Rule{S:WhileStepAbort}
{  k < \llbracket x - y \rrbracket_H
\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
}
{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }

\and \Rule{S:If}
{  (H, S_i) \Downarrow_{(q, q')} X
\\ i \in \{1, 2\}
}
{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_{M_c^i \cdot (q, q')} X }

@@

\end{mathpar}

\section{Soundness theorem}

@@

\begin{proof} By induction on the big step semantics.

@@

\item Case {\sc S:Skip}.
  We have $(H, \code{skip}) \Downarrow_{M_s} H$.
  By assumption, $H \models \Gamma$. \\
  Since $\Phi_H(Q + M_s) = \Phi_H(Q) + M_s$,
  \begin{itemize}[topsep=0pt]
  \item
    if $M_s \ge 0$ we have,
      $M_s \le \Phi_H(Q + M_s)$ and
      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge M_s - 0$;
  \item
    if $M_s < 0$ we have,
      $0 \le \Phi_H(Q + M_s)$ and
      $\Phi_H(Q + M_s) - \Phi_H(Q) \ge 0 - (-M_s)$.
  \end{itemize}

@@

>>>>

\usepackage{amsthm}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

@@

\section{Introduction}

After functional correctness, quantitative properties are the
second thing one wants to know when writing a program.
For example, an asymptotic runtime bound on a
sorting algorithm tells us what efficiency we can expect
of it when running on very large arrays.  But, what if we
always run it on arrays of size between 5 and 10? In that
case, the asymptotic is not relevant and we need a more
precise bound on the runtime.  It is then a good
software engineering practice to profile the algorithm on
some examples.

In between asymptotic bounds and profiling there are
precise bounds, for instance, one can prove that Knuth's Quicksort might
take at most $20N^2$ operations of a MIX machine when
called on an array of size $N$.  If you have this information,
no profiling is needed to compare the algorithms and you
have a formal guarantee on the behavior of the program. But what if you do not
have a MIX machine, or you tweaked Knuth's algorithm
a bit?  Then you are good for another long an tedious
proof.  We advocate that, it is possible to design a practical
system that is able to infer precise worst case bounds
without any user interaction.

This work presents a resource oblivious system to get
linear resource bounds on imperative programs with
integers.  Despite its apparent simplicity it is able to
reproduce multiple results from the literature which
were obtained using methods based on abstract
interpretation.  In stark contrast with abstract-interpretation
based methods, our tool does not require any
fixpoint computations on an abstract domain to obtain
loop invariants.  The mechanism we designed, which
is based on amortized analysis, is able to leverage a local
information to obtain global invariants.  Additionally, the
approach we advocate is able to scale to full blown
imperative languages including constructs like functions
calls, \code{break} and \code{return}.

The analysis is proved sound with respect to
the operational semantics instrumented to track resource
consumption.  It is implemented as a prototype
in OCaml and available on demand.  We test it on multiple
algorithms from the literature and on custom tests, it
is also planned to extend it to handle the full C programming
language.

\section{Overview and Examples}

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized
analysis.  Say that a program $p$ evaluates on a heap $H$
by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
The basic idea of amortized analysis is to define a potential function
$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.

Since imperative programs can be composed we must enrich
the judgement $(\_, \_) \Downarrow_{\_}$ to talk about the heap
resulting from the program execution $H'$.  So we now write
$(p, H) \Downarrow_n H'$ to say that $H'$ results from $p$'s
execution.  We now use two potential functions, one
that applies before the execution, and one that applies
after.  The two functions must respect the natural relation
$\Phi(H) \ge n + \Phi'(H')$.  Intuitively speaking, $\Phi$ must
give enough \emph{potential} to, both, finish the computation
and pay for $\Phi'$ on the resulting heap. That way, if $(H, p_1) \Downarrow_n H'$
and $(H', p_2) \Downarrow_m H''$, we get $\Phi(H) \ge
n + \Phi'(H')$ and $\Phi'(H') \ge m + \Phi''(H'')$, which
can be composed to give $\Phi(H) \ge (n + m) + \Phi''(H'')$.
What we have observed is that, if we define $\Phi \vdash p \dashv \Phi'$
to be
$$
\forall H\,n\,H',~(H, p)\Downarrow_n H' \implies \Phi(H) \ge
n + \Phi'(H'),
$$
then we get the following familiar looking rule:
$$
\RuleNolabel
{\Phi \vdash p_1 \dashv \Phi' \\ \Phi' \vdash p_2 \dashv \Phi''}
{\Phi \vdash p_1; p_2 \dashv \Phi''}
.
$$
Similarly, other language constructs contrain the potential
functions to respect more rules that look very similar to
Hoare logic or effect systems rules.  By fixing the shape of
a potential function $\Phi$, it becomes possible to leverage
existing tools to help finding a pair of potential functions
for an arbitrary program.  This is explained in further details
in the following sections.

We say that the analysis is amortized because the use of a
potential function $\Phi$ gives the flexibility to amortize some
operations.  For example, if heaps are simple mappings from
$\{\code x, \code y\}$ to natural numbers, we have the following
judgement (provided that $H~\code x \ge H~\code y \ge 0$ initially)
$$
(\Phi := \lambda H.~2\cdot(H~\code x) + 10)
\vdash \code x = \code x - \code y \dashv
(\Phi' := \lambda H'.~2\cdot(H'~\code x) + 2\cdot(H'~\code y) + 10).
$$
This must be understood as a motion of potential from $\code x$
to $\code y$. If some resource is consumed $\code y$ times afterwards, we
can now use $\code y$'s potential to pay for it. But interestingly, the
initial potential function $\Phi$ only talks about~$\code x$: the potential
function allowed us to impart the cost of the subsequent resource
consumptions on the variable~$\code x$.
This flexibility combined with intelligent decisions on potential
motions allow our tool to discover linear invariants for surprisingly
convoluted programs.

\subsection{Motivating Examples}

We now present a representative set of examples that can be
handled by our tool.  The bounds that are given are for the number
of ``base statements'' executed by the program, that is, the number
of assertions, skip and assignment statements executed.  The
bounds are given as literal expressions where a variable name {\tt x}
represents the initial value of this variable.

\paragraph{Simple loops.} The simplest loop scheme that is in
use in typical imperative programs is counting upwards or
downwards to a given bound N.  The two example programs
are given below, on their right is the bound inferred by our analysis.
\begin{center}
\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
\lstinline|while x > N ( x = x - 1 )| & $|{\tt N} -  {\tt x}|$ \\
\lstinline|while x < N ( x = x + 1 )| & $|{\tt N} -  {\tt x}|$ \\
\end{tabular}
\end{center}

It is also frequent to iterate every $n$ elements in an array, also
in this simple case our analysis works without any problem.  Note
however that if we increment using a variable, it might very well be
possible that the program diverges!  In this situation no worst
case bound can be given and the analysis correctly fails.
\begin{center}
\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
\lstinline|while x <= -2 ( x = x + 2 )| & $| {\tt x} | / 2$ \\
\lstinline|while x <= z - y ( x = x + y )| & Infeasible \\
\lstinline|assert y > 0; while x <= z - y ( x = x + y )| & $1 + |{\tt x} - {\tt z}|$
\end{tabular}
\end{center}
In the last example, the constant 1 is added to the loop bound
to account for the initial \texttt{assert} that consumes one
step.  Also, since the bound is for the worst case, it is tight
only when {\tt y} is 1.

\paragraph{In the literature.}
The two examples below are taken from Gulwani's paper~\cite{Gulwani-speed}.
They are both handled by the Speed tool but require inference of a so-called
disjunctive invariant.  In the abstract interpretation community, these
invariants are known to be notoriously difficult to handle.

\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
while x < n
  if y < m    y = y + 1
  else        x = x + 1
\end{lstlisting}
$$
|{\tt m} - {\tt y}| + |{\tt n} - {\tt x}|
$$
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
while x < n
  if z > x    x = x + 1
  else        z = z + 1
\end{lstlisting}
$$
|{\tt n} - {\tt x}| + |{\tt n} - {\tt z}|
$$
\end{minipage}

\hspace{1ex}

While the first
example is simple to understand, the second one is much more
tricky and trying to understand how it actually works is quite challenging.
However, with the amortized analysis in mind, using the potential transfer
reasonning, it is almost trivial to prove the termination.  While the Speed
tool has to find a fairly involved invariant for the loop, our tool is simply
reasoning locally and works without any special trick.
%
Our analysis handles without problems all the other examples
in~\cite{Gulwani-speed} that have a linear bound.  Here is
a quick review of some of them together with the bounds we
inferred.

\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n
  while x < n
    x = x + 1
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
$$
\end{minipage}
\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n (x = x + 1);
while x < n (x = x + 1)
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
$$
\end{minipage}
\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n
  ( while y < m (y = y+1)
  ; x = x + 1 )
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
+|{\tt y} - {\tt m}|
$$
\end{minipage}

\paragraph{Amortized reasonning.}
The two examples below show the power of amortized reasoning.
Because they essentially rely on an amortized reasonning we believe
that our tool is the only one able to deal with them.

\begin{minipage}[b]{0.45\textwidth}
\begin{lstlisting}
assert y >= 1;
while x + y <= N
  ( x = x + y
  ; z = y; z = z + 4
  ; while z > 0 ( z = z - 1 ) )
\end{lstlisting}
$$
1 + 8\,|{\tt N} - {\tt x}|
$$
\end{minipage}
\begin{minipage}[b]{0.45\textwidth}
\begin{lstlisting}
while x > 0
  ( x = x - 1
  ; if ?
      while s > 0 ( s = s - 1 )
    else s = s + 1 )
\end{lstlisting}
$$
3\,|{\tt x}| + |{\tt s}|
$$
\end{minipage}

\hspace{1ex}

Maybe surprisingly, the two examples have linear bounds.  In the first one, the
worst case is when {\tt y} is 1, in this situation the inner loop will perform 5 times
$|{\tt N} - {\tt x}|$ iterations. Indeed, for each decrement of {\tt x}, {\tt z} is set
to 5 and iterated over.  The second example is classic in the amortized analysis
community, it is the original example from Tarjan's paper~\cite{Tarjan-amort} that
non deterministically pushes one element on a stack or pops many elements from
that stack.  In the example source code, the stack depth is represented by the
variable {\tt s} and the number of iterations is ruled by the initial value of {\tt x}.

\section{Syntax}

Let us now start the formal part of this work.  We begin by defining
a proper syntax for the language we will be using later on.  This is
a more algebraic version of the surface syntax used in the previous
examples.
%
\begin{mathpar}

S := \code{skip}
\mid \code{assert}(C)
\mid x \gets E
\mid \code{while}(C)~S
\mid \code{if}(C)~S~\code{else}~S
\mid S;S

@@

E := x \mid x \pm y
\end{mathpar}

Note that expressions are kept extremely simple, this will help
greatly when designing the analysis.  Moreover, programs
containing fancier expressions can easily be compiled to this
restricted form.  For example $a \gets b + c - d$ is semantically
equivalent to $a \gets b; a \gets a + c; a \gets a - d$ which
fits in the above syntax.

The syntax of conditions, noted $C$, is left unspecified
for the moment.  We will describe it later when the practical
implementation details are provided.  The only semantic tool
we need to deal with this unknown is an evaluation
function $\llbracket \_ \rrbracket_{\_}$ such that, for any
condition $C$ and any heap $H$ we have $\llbracket
C \rrbracket_H \in \{ \top, \bot \}$.

Another peculiarity is that there are no integer constants in the
syntax.  Once again this is for the sake of simplicity.  In practice
the concrete syntax recognizes ``variables'' like {\tt -4} or
{\tt 60} as read only variables.  Any heap used later in the
operational semantics will map these variables to their natural
value.

\section{Cost Semantics}

The semantics for the syntax defined above needs to account
for resource consumption.  To account for all kinds of resources
we leave it as a parameter of the semantics to be specified
by the user.  This parameter, called a metric, is a tuple of rational
numbers with the following form
$$
(
M_k, M_s, M_a,
M_w^1, M_w^2, M_w^3,
M_c^1, M_c^2,
M_s^1, M_s^2
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that some resource is released by the program.

The general form of an evaluation judgement is
$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
presented in the overview section appart  from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid with by taking
$(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.  This injection
is used implicitely in the semantics.

\begin{figure}[ht!]
\begin{mathpar}

\Rule{S:Abort}
{ }
{ (H, S) \Downarrow_{(0,0)} \circ }

\and \Rule{S:Skip}
{ }
{ (H, \code{skip}) \Downarrow_{M_k} H }

@@

\and \Rule{S:Assert}
{ H \models C }
{ (H, \code{assert}(C)) \Downarrow_{M_a} H }

\and \Rule{S:WhileStop}
{ \llbracket C \rrbracket_H = \bot }
{ (H, \code{while}(C)~S) \Downarrow_{M_w^3} H }

\and \Rule{S:WhileAbort}
{  k < \llbracket x - y \rrbracket_H
\\ (H, S) \Downarrow_{(q_1, q'_1)} \circ
}
{ (H, \code{while}(k < x - y)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) } \circ }

\and \Rule{S:WhileStep}
{  \llbracket C \rrbracket_H = \top
\\ (H, S) \Downarrow_{(q_1, q'_1)} H'
\\ (H', \code{while}(C)~S) \Downarrow_{(q_2, q'_2)} X
}
{ (H, \code{while}(C)~S) \Downarrow_{M_w^1 \cdot (q_1, q'_1) \cdot M_w^2 \cdot (q_2, q'_2)} X }

\and \Rule{S:IfThen}
{  (H, S_1) \Downarrow_{(q, q')} X
\\ \llbracket C \rrbracket_H = \top
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }

\and \Rule{S:IfElse}
{  (H, S_2) \Downarrow_{(q, q')} X
\\ \llbracket C \rrbracket_H = \bot
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }

@@

\end{mathpar}
\caption{Operational semantics of statements}
\label{fig:opsem}
\end{figure}

Heaps manipulated by the operational semantics judgements are
simple maps from variable names to integers.  We designate by
$H[x \mapsto v]$ the heap $H$ updated to map the variable name
$x$ to the number $v$.
%
The evaluation of expressions is trivial using heaps.
\begin{mathpar}
\llbracket x \rrbracket_H = H(x)
\and
\llbracket x \pm y \rrbracket_H = H(x) \pm H(y)
\end{mathpar}

In the evaluation judgement $(H, S) \Downarrow_{(q, q')} X$, the result $X$ can
be either a new heap, which means the execution terminated successfully
or it can be the \emph{busy} result $\circ$.  This indicates an early
termination, it can be triggered at any point using the rule {\sc S:Abort}
and bubbles up through the execution context.
Sometimes there is no other choice than to use this latter rule, for
example when an assertion fails.  In that case the side condition of
{\sc S:Assert} is not met and the only rule that can possibly be applied
is {\sc S:Abort}.

The busy result is also useful to handle diverging programs with big-step
judgements.  In case a program does not terminate, there is no heap $H'$
such that $(H, S)\Downarrow_{(q, q')} H'$.  The only possible way to
conclude a big-step judgement is to use the {\sc S:Abort} rule at some
point in the computation.  One alternative to this technique is to use
coinductive semantics~\cite{Leroy-coind}, however, using the busy
result is precise enough in our setting of quantitative resource analysis and
allows to have only one soundness proof for both unsafe and
non-terminating programs.  Indeed, stating a result for
$(H, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
programs, and partial computations at the same time.

\section{Proof System}

We define in this section an abstract version of the proof system that
we will use to derive bounds in our practical implementation.

\subsection{Linear Potential Functions}

To get a practical and automated implementation of amortized analysis
as presented in the overview section we need the fix the shape of
potential functions.  The general form of a potential function is as follows
$$
\Phi(H) = q_0 + \sum_{a \prec b} q_{ab}\,|H(a) - H(b)|.
$$
Where $a$ and $b$ range over the set of all program variables $V$ and
$(q_{ab})$ and $q_0$ are non-negative rational numbers.  The relation
$\prec$ is an arbitrary complete order on $V$ that is used only to
avoid considering the two equal quantities $|H(a) - H(b)|$ and
$|H(b) - H(a)|$ in the above sum.

To simplify the references to the $q$ coefficients later we introduce
an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
\{(a, b) \mid a \prec b\}$.  Each index $i$ corresponds to a \emph{base
function} $f_i$ in the potential function: $0$ corresponds to the constant
function $\_ \mapsto 1$, and $(a,b)$ corresponds to $H \mapsto
|H(a) - H(b)|$.  Using these notations we can rewrite the above
equality as
$$
\Phi(H) = \sum_{i \in I} q_i f_i(H).
$$
From now on, we will write $ab$ to denote the index $(a,b)$.

\subsection{Proof Rules}

\begin{figure}[ht!]
\begin{mathpar}

\Rule{Q:Skip}
{}
{ \Gamma; Q + M_k \vdash \code{skip} \dashv Q; \Gamma }

\and \Rule{Q:Assert}
{}
{ \Gamma; Q + M_a \vdash \code{assert}(C) \dashv Q; \Gamma, C }

\and \Rule{Q:While}
{ \Gamma, C ; Q - M_w^1 \vdash S \dashv M_w^2 + Q; \Gamma }
{ \Gamma; Q \vdash \code{while}(C)~S \dashv Q - M_w^3; \Gamma, \neg C }

\and \Rule{Q:Set}
{ \forall u,~q_{yu} = q'_{xu} + q'_{yu}
\\ q'_{xy} \in \mathbb Q_0^+
}
{ \Gamma[x/y]; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }

\and \Rule{Q:Incr}
{ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
\\ \textstyle q'_y = q_y
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma[x/x \pm y]; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
}

\and \Rule{Q:If}
{ \Gamma, C; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma, \neg C; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
}
{ \Gamma; Q \vdash \code{if} (C)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }

\and\Rule{Q:Seq}
{  \Gamma; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }

\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_\Gamma Q_2
\\ \Gamma_2; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
\\ Q'_2 \succeq_\Gamma Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ \Gamma_1; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }

\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |x - y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |x - y| \le u_{xy} \}
%\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
\\ \forall i \in \mathcal U,~ q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L,~ q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L,~ q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }

\end{mathpar}
\caption{Abstract proof rules of the quantitative analysis}
\label{fig:proof}
\end{figure}

The deductive system defined in~\pref{fig:proof} derives
judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
Its semantics are formally described in the soundness section but we
will give an intuition of its meaning here.

\paragraph{Quantitative part.}
The judgement is really two parts, one logical part, and one quantitative
part.  The quantitative part is, as shown in the overview section, a
representation of two potential functions $\Phi$ and $\Phi'$ using
two maps from indices to non-negative rational numbers
$Q, Q' : I \rightarrow \mathbb Q_0^+$, these maps are called
\emph{quantitative contexts}.
We note $q_i$ for $Q(i)$.  Since $Q$ specifies a value for each
index in $I$ it is clear that it defines a unique linear potential function
as described in the previous section. We define $H \mapsto \Phi_H(Q)$
to be the linear potential function canonically associated to $Q$.

The rules of~\pref{fig:proof} are designed to enforce a semantic
soundness similar to the one exposed in the overview section.  To
do this, the rules impose contraints on $Q$ and $Q'$.  To make the
presentation lighter, we use the following  set of conventions.
First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
and $\forall i \neq 0,~q'_i = q_i$.  Second, for any index $i$ such that
the relation between $q_i$ and $q'_i$ is not specified as a side condition
of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
0 is a program variable symbolizing the constant 0.

\paragraph{Logical part.}
The second part of judgements maintains an abstract program state.
This should be seen as a very weak Hoare logic.  It has to be weak because
we want to automate it trivially and it can be weak because we only use
it to get very local and simple knowledge.  The specifics of logical assertions
in $\Gamma$ are left to the implementation section.
%
The purpose of this logical state is to help the rules
make some sound decisions when relating coefficients of quantitative
contexts.

Syntactically $\Gamma$ is a list of program conditions.  Because of the
rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
to be closed by negation.  Semantically we define the validity
$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
%
The logical state
is maintained using regular Hoare style rules.
The semantic entailment on logical states is noted $\Gamma \models \Delta$,
its meaning is that $\forall H,~ H\models\Gamma \implies H\models\Delta$.

%\paragraph{Explanation of selected rules.}

\section{Soundness Theorems}

The soundness of a judgement relates the cost semantics with the potential
functions derived using our proof system.  Depending on the outcome of
the semantics, we give different soundness statements.
But, before tackling the issue of soundness let us remind and define some
convenient notations.
\begin{itemize}
\item $H \mapsto \Phi_H(Q)$ for $Q$ a quantitative context, is the
unique linear potential function that has $Q(i)$ as a coefficient for the
base function $f_i$.
\item $H \models \Gamma$ is true if and only if all the assertions in
$\Gamma$ are true on the heap $H$.
\end{itemize}

Ideally we would
like our judgements to be meaningful even in the case of a program
crash.  More precisely, we want the first potential function to be enough
to pay for the execution of the program up to the crash point.  This is
formally stated in the following theorem.

\begin{theorem}[Busy case]
For any $S$, $H$, $q$, $q'$, $\Gamma$, $\Gamma'$, $Q$ and $Q'$
such that $H \models \Gamma$, $\Gamma; Q \vdash S \dashv Q'; \Gamma'$,
  and $(H, S) \Downarrow_{(q,q')} \circ$,
we have $q \le \Phi_H(Q)$.
\end{theorem}
\begin{proof}
Simple application of the terminating case proved afterwards.
\end{proof}

In order to prove the harder case for terminating programs, need to first
focus on the only structural rule of the proof system: {\sc Q:Weak}.  This
rule can be applied anywhere in the derivation, it is not syntax directed.
To prove the soundness of this rule, we need a lemma that gives a
semantic explanation of the $\succeq_\Gamma$ relation.

\begin{lemma}[Relax]
If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
we have $\Phi_H(Q') \ge \Phi_H(Q)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_H(Q)
  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
 &\ge \Phi_H(Q).
\end{align*}
Since $H \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L,~ l_i \le f_i(H)$ and
$\forall i\in\mathcal U,~ f_i(H)\le u_i$.
These two inequalities justify the last step.
\end{proof}

Using this lemma, we can tackle the hardest theorem of soundness.  This theorem
needs a strong statement to make the proof by induction go through and defines
the complete meaning of a judgement $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
In a similar fashion than the busy case, the initial quantitative context is enough
to perform the whole computation: $q \le \Phi_H(Q)$.  But additionally, the
two quantitative contexts are also related to the resource consumption $q - q'$
of the piece of program executed.  This second requirement exactly matches
the semantic condition described in the overview section: it says that the
initial potential $\Phi_H(Q)$ is enough to both pay for the resource
consumption of the program $q - q'$ and pay for the potential of the resulting
heap $\Phi_{H'}(Q')$.

The proof of this result if very technical but basically goes through by induction
on the big-step evaluation.  Only a representative subset of all the cases
is present in the following proof.

@@

\begin{proof} By nested induction on the big step semantics and on the
judgement derivation (to account for the structural rule {\sc Q:Weak}).

@@

\item Case {\sc S:Skip}.
  We have $(H, \code{skip}) \Downarrow_{M_k} H$.
  By assumption, $H \models \Gamma$. \\
  Since $\Phi_H(Q + M_k) = \Phi_H(Q) + M_k$,
  \begin{itemize}[topsep=0pt]
  \item
    if $M_k \ge 0$ we have,
      $M_k \le \Phi_H(Q + M_k)$ and
      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge M_k - 0$;
  \item
    if $M_k < 0$ we have,
      $0 \le \Phi_H(Q + M_k)$ and
      $\Phi_H(Q + M_k) - \Phi_H(Q) \ge 0 - (-M_k)$.
  \end{itemize}

@@

\section{Implementation}

\subsection{Logical Annotations Generation}
